---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# VariantStoryBench

[![R-CMD-check](https://github.com/RGenomicsETL/VariantStoryBench/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/RGenomicsETL/VariantStoryBench/actions/workflows/R-CMD-check.yaml)

VariantStoryBench is an R package for reproducible rare-disease variant
reanalysis benchmarks. It validates compact result manifests, calculates
ranking and candidate-burden metrics, and provides an S7 adapter contract for
running declared external engines.

## Install

```r
remotes::install_github("RGenomicsETL/VariantStoryBench")
```

## Metrics

```{r, eval=FALSE}
library(VariantStoryBench)

results <- bench_read_results(
  system.file("extdata", "synthetic-results.csv", package = "VariantStoryBench")
)
bench_metrics(results)
```

`causal_rank = NA` represents a no-causal case. Causal recall is calculated
only over cases with a declared causal rank; no-causal cases remain in candidate
burden metrics.

## External engines

`BenchCommandAdapter` uses `blit` to run an executable with declared, quoted
arguments. `{input}` and `{output}` are substituted before execution.
`BenchRunner` is the S7 contract accepted by consumers of adapters.

```{r, eval=FALSE}
adapter <- BenchCommandAdapter(
  executable = "my-prioritizer",
  arguments = c("--input", "{input}", "--output", "{output}")
)

run <- bench_execute(adapter, "case.vcf", "benchmark-output/case.csv")
stopifnot(run@status == 0L)
```

## Pipeline

The repository contains a [`targets`](https://docs.ropensci.org/targets/)
pipeline over the packaged synthetic manifest.

```r
# From the package source directory:
targets::tar_make(callr_function = NULL)
```

The initial fixture is synthetic. Add public or redistributable fixtures with
their source, version and licence recorded alongside the data.

## Licence

GPL-3. See `DESCRIPTION`.
