---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# VariantStoryBench

[![R-CMD-check](https://github.com/RGenomicsETL/VariantStoryBench/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/RGenomicsETL/VariantStoryBench/actions/workflows/R-CMD-check.yaml)

VariantStoryBench generates small sealed GRCh38 benchmark inputs and evaluates
declared rare-disease prioritisation results against separate evaluator truth.
It does not implement a prioritiser, generic command runner, workflow
framework, or source ingestion/provenance system.

[ARCHITECTURE.md](ARCHITECTURE.md) states the current boundary and
[benchmark-contract.md](benchmark-contract.md) defines executable relations,
coordinate conventions, and metric denominators.

## Install

```r
remotes::install_github("RGenomicsETL/VariantStoryBench")
```

## Generate sealed inputs and evaluator truth

The generator writes causal singleton, trio, symbolic-CNV, and phenotyped
confirmed-negative VCFs with `vcfppR`.
The engine-facing bundle contains only VCFs, GRCh38 generations, persons with
exact VCF sample IDs, family presentation, and clinical documents. `persons`
has `case_id`, `person_id`, `vcf_sample_id`, `is_proband`, `sex`, and `affected`;
`relationships` uses `case_id`, `person_id`, `relative_id`, and
`relationship`. The evaluator-only bundle contains causal answers and latent
observations. No reference result or oracle is supplied.

The generated records use Ensembl GRCh38 primary-assembly contig names and
these verified reference alleles:

| VCF locus | REF | ALT |
|:--|:--|:--|
| `1:100000` | `C` | `T` |
| `1:100100-100102` | `TGC` | `TC` (canonical `TG>T`) |
| `2:199999-200000` | `TG` | `T` |
| `2:200100` | `T` | `A` |
| `2:200200` | `A` | `C` |
| `2:200300` | `T` | `C` |
| `7:549997` | `T` | `<DEL>` (`END=559997`) |

These micro-coordinates were checked against Ensembl release 116 GRCh38
primary assembly. The reference FASTA is not bundled with the package.

```{r generation}
library(VariantStoryBench)

bundle <- bench_generate_micro_cohort(tempfile("variantstorybench-micro-"))
paste(names(bundle$engine_input), collapse = ", ")
paste(names(bundle$evaluator_truth), collapse = ", ")
bundle$engine_input$generations[, c("generation_id", "assembly", "case_design")]
```

A text provider receives only a case ID, HPO/context presentation, and family
relationships. It must return valid source text for every case, including an
empty phenotype plan, and zero-based half-open spans for every presented HPO
observation. The CNV note therefore has zero observations. It never receives
causal truth.

```{r, eval=FALSE}
provider <- function(presentation) {
  # Call a configured gpt-5.3-spark or Rbebelm provider outside this package.
  terms <- presentation$phenotype_plan
  source_text <- if (nrow(terms)) {
    paste(terms$hpo_id, collapse = "; ")
  } else {
    "No phenotype terms were supplied."
  }
  starts <- vapply(terms$hpo_id, function(id) {
    regexpr(id, source_text, fixed = TRUE)[[1L]] - 1L
  }, integer(1))
  list(source_text = source_text, spans = data.frame(
    hpo_id = terms$hpo_id,
    context_status = terms$context_status,
    start_offset = starts,
    end_offset = starts + nchar(terms$hpo_id)
  ))
}
```

Each source document has `case_id`, `document_id`, and `source_text`; there is
one document and one proband for every case. Generated evaluator observations
wrap `case_id`, `person_id`, and `observation_id` around the unchanged canonical
`ducksemantics` fields: `document_id`, `hpo_id`, exact source span/text,
`context_status`, method, provider provenance, confidence, and accepted status.
The wrapper prevents a relative's finding from being scored as the proband's;
document text/span semantics remain owned by ducksemantics. Negated findings
are `absent/negated`.

```{r hpo}
bench_hpo_metrics(
  bundle$evaluator_truth$documents,
  bundle$evaluator_truth$hpo_observations,
  bundle$evaluator_truth$hpo_observations
)[, c("term_recall", "context_recall", "span_recall")]
```

## Declared engine results

There is currently no bundled engine adapter. A caller may run a concrete
sealed adapter outside the package, passing only `bundle$engine_input`, then
supply declared `runs`, `evaluations`, and `candidates` to evaluation with
`bundle$evaluator_truth`.

```{r, eval=FALSE}
metrics <- bench_evaluate_micro_cohort(
  bundle$evaluator_truth,
  runs,
  evaluations,
  candidates
)
```

Ranks are unique and contiguous per run/case/target, with score ties broken by
candidate ID. Failed, skipped, and unsupported known-causal units remain
ranking misses. Source spans are strictly nonempty zero-based half-open
intervals.

A frequency-capable adapter may additionally emit the narrow typed
case-frequency relation validated by `bench_validate_frequency_observations()`.
The validator keeps absent rows, unlocalized shards, provider conflicts,
filtering, missing values, AC=0/AN>0, and AN=0 distinct; it requires lowercase
16-character hexadecimal VariantKey text, checks AC/AN/AF and FAF bounds and
missingness, and never assigns PM2. The broad provider pack is not a benchmark output.

## Sealed scientific truth relations

The evaluator-only bundle retains source/canonical `allele_truth`, person-grain
`genotype_truth`, explicit `causal_allele_truth`, `inheritance_truth`, and
`cnv_truth` as sealed relations.
`bench_allele_transport_metrics()` compares source and canonical GRCh38
geometry plus source admission by record/ALT ordinal.
`bench_genotype_transport_metrics()` compares GT, GQ, DP, ALT-relative copy
count, ploidy, phase/phase set, and ALT-relative call state without applying a
GQ threshold. `causal_allele_truth` separately marks source `reference` versus one
exact `alternate` ordinal: the singleton causal answer intentionally targets
REF, while the confirmed-negative case has a called ALT and no causal row.
Thus REF/ALT and call state never stand in for pathogenic/benign orientation.
These truth relations are not sent to an engine or text provider.

The symbolic deletion is a real `cnv` case with GRCh38, type, BED interval,
source `unsupported_symbolic` admission, and separate `routed`/`XCNV` routing
truth. VCF `POS=549997` is the one-based padding/base-before-event
coordinate, while `END=559997` is the inclusive last affected base. Excluding
the padding base gives the affected BED half-open interval `[549997, 559997)`:
use VCF `POS` as the BED start and `END` as the exclusive end. This is a
10,000 bp affected interval. The VCF itself remains unchanged. Exact GQ and DP
are genotype transport truth, not evidence-policy thresholds.

The capability relation explicitly marks realistic exome/genome simulation,
related ancestry/admixture, multiplex/consanguinity, general CNV/SV, and novel
gene-disease historical holdouts as unsupported.

## Temporal status

Release-pinned temporal selection and source-snapshot anti-join proof are not
implemented. `bench_classification_diff()` remains available for a
caller-provided fixed-policy comparison of two RClinVarbitration decisions;
this package does not accept generic anti-join counts as proof.

## Pipeline

[`targets`](https://docs.ropensci.org/targets/) generates separate engine-input
and evaluator-truth targets and validates their contracts. It does not invent
engine output.

```r
targets::tar_make(callr_function = NULL)
```

## Licence

GPL-3. See `DESCRIPTION`.
